---
title: "AlphaFold Analysis"
author: Bayah Essayem (A17303992)
format: pdf
date: February 11 2025
toc: TRUE
---

Here we analyze our AlphaFold structure prediction models. The input directory/folder comes from the ColabFold server:

# Change this for YOUR result dir name
```{r}
results_dir <- "hivpr_monomer_94b5b.result/" 

```

# File names for all PDB models

```{r}

  
  
  
  
  
  
  pdb_files <- list.files(path="hivpr_monomer_94b5b.result/hivpr_monomer_94b5b/",
                        pattern="*.pdb",
                        full.names = TRUE)


# Print our PDB file names
basename(pdb_files)
```

I will use the Bio3D package for analysis

```{r}
library(bio3d)
```

Align and superpose

```{r}
# Read all data from Models and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile ="msa")
pdbs

```

A quick PCA

```{r}
pc <- pca(pdbs)
plot(pc)
```

# RMSD analysis
RMSD (root, mean, square, of the distance) is a common measure of structural distance used in structural biology.

```{r}
rd <- rmsd(pdbs, fit = T)
rd
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

