---
title: "class10 (cont.)"
author: "Bayah Essayem (A17303992)"
date: February 11, 202
format: pdf
toc: TRUE
---

Here we run through a complete "pipeline" of structure analysis that begins with a single sequence identifier and ends in a PCA analysis.

```{r}
library(bio3d)
```

Step 1. 
Retrieve a sequence for the protein we are interested in. We will take ADK "1ake_A

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

## Step 2.
Run a BLAST search of the PDB for all related sequences to our input 'aa' 

```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```

## Step 3. Download all structures
These are our "top hits" i.e. all the structures in the PDB database related to our input sequence.

```{r}
hits$pdb.id
```

# Download related PDB files
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split = TRUE, gzip = TRUE)
```

## Step 4. Align and superpose
# Align releated PDBs

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Step 5. PCA

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 

```{r}
pc <- pca(pdbs)
plot(pc)
```

Let's make a trajectory (or movie) of the main conformational changes captured by PC1. We will use the 'mktrj()' function for this...

```{r}
mktrj(pc, file = "pca_result.pdb")
```


Back of the envelope comparison of the PDB size to UniProt
PDB: 231029, UniProt size: 2531029

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot * 100
```

# Calculate RMSD

```{r}
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```


